---
title: "Data Preparation"
subtitle: "Neighbourhood Transport Environment Typology"
author: "Laurie Platt"
date: last-modified
date-format: "[Last updated ] DD MMMM, YYYY"
format:
  html:
    css: resources/styles.css
    code-tools: true
    self-contained: false
    code-fold: true
    toc: true
    toc-depth: 5
    toc-expand: 1
    number-sections: true
    fig-cap-location: bottom
    fig-align: left
    code-links:
      - text: scc-pi/transport-env-typology
        icon: github
        href: https://github.com/scc-pi/transport-env-typology
execute:
  warning: false
  message: false
---

We're using the {targets}^[{targets} user manual ([books.ropensci.org/targets/targets.html](https://books.ropensci.org/targets/targets.html))] R package to help manage our machine learning (ML) pipeline. 

```{r setup}

# Packages
library(tidyverse)
library(here)
library(targets)
library(gt)
library(gtExtras)
library(sf)
library(leaflet)
library(osdatahub)
library(readxl)

# Sources all R/*.R same as _targets.R
tar_source()       

# Pull any pre-computed targets we need
tar_load(sf_city)
tar_load(sf_lsoa)

# ggplot theme
set_theme(theme_minimal())
```

# Assemble datasets  

## Ingest

```{r}
#| label: tbl-ingest
#| tbl-cap: "Data ingested (via `_targets.R` & `R\\ingest.R`)"
#| column: page-right

path_transport_data_xls <- here("data","sheffield_transport_datasets.xlsx")

# Read a list of our raw datasets from a spreadsheet
df_ingested <- read_xlsx(
  path_transport_data_xls, 
  sheet = "Raw datasets", 
  skip = 3
)

# List the raw datasets ingested
df_ingested |>
  mutate(Row = row_number(), .before = 1) |>
  my_gt() |>
  fmt(
      columns = "Row",
      fns = function(x) {format(round(x, 0), nsmall = 0, big.mark = ",")}
    ) |>
  cols_label(Row = "") |>
  opt_css("
    td:nth-child(6) {
      overflow-wrap: anywhere;
      word-break: break-word;
    }
  ")
```

## Process

```{r}
#| label: tbl-process
#| tbl-cap: "Data processed (via `_targets.R` & `R\\process.R`)"
#| column: page-right

# Read a list of our processed data from the same spreadsheet
df_processed <- read_xlsx(
  path_transport_data_xls, 
  sheet = "LSOA indicators", 
  skip = 3
)

# List the processed datasets
df_processed |>
  mutate(Row = row_number(), .before = 1) |>
  my_gt() |>
  fmt(
      columns = "Row",
      fns = function(x) {format(round(x, 0), nsmall = 0, big.mark = ",")}
    ) |>
  cols_label(Row = "")
```

Note: `sf_lsoa_imd` ingested at LSOA level courtesy of MHCLG (see @tbl-ingest).

## Spatial data  

BNG is our default CRS. Some visualisation tools such as leaflet require WGS84, but transformation for this will only be done at the point of visualisation.  

# Handle missing values

# Standardise variables

# Explore distributions
  
  
```{r}
#| label: city-boundary

make_basemap() 
```
  
# Examine correlations

