---
title: "Data Preparation"
subtitle: "Neighbourhood Transport Environment Typology"
author: "Laurie Platt"
date: last-modified
date-format: "[Last updated ] DD MMMM, YYYY"
format:
  html:
    css: resources/styles.css
    code-tools: true
    self-contained: false
    code-fold: true
    toc: true
    toc-depth: 5
    toc-expand: 1
    number-sections: true
    fig-cap-location: bottom
    fig-align: left
    code-links:
      - text: scc-pi/transport-env-typology
        icon: github
        href: https://github.com/scc-pi/transport-env-typology
execute:
  warning: false
  message: false
---

We're using the {targets}^[{targets} user manual ([books.ropensci.org/targets/targets.html](https://books.ropensci.org/targets/targets.html))] R package to help manage our machine learning (ML) pipeline. 

```{r setup}

# Packages
library(tidyverse)
library(here)
library(targets)
library(gt)
library(gtExtras)
library(sf)
library(leaflet)
library(osdatahub)
library(readxl)

# Sources all R/*.R same as _targets.R
tar_source()       

# ggplot theme
set_theme(theme_minimal())
```

# Assemble datasets  

```{r}
#| label: tbl-ingest
#| tbl-cap: "Data ingested (via `_targets.R` & `R\\ingest.R`)"
#| column: page-right

path_transport_data_xls <- here("data","sheffield_transport_datasets.xlsx")

# Read a list of our raw datasets from a spreadsheet
df_ingested <- read_xlsx(
  path_transport_data_xls, 
  sheet = "Raw datasets", 
  skip = 3
)

# List the raw datasets ingested
df_ingested |>
  mutate(Row = row_number(), .before = 1) |>
  my_gt() |>
  fmt(
      columns = "Row",
      fns = function(x) {format(round(x, 0), nsmall = 0, big.mark = ",")}
    ) |>
  cols_label(Row = "") |>
  opt_css("
    td:nth-child(6) {
      overflow-wrap: anywhere;
      word-break: break-word;
    }
  ")
```

::: {.callout-note}
## Decision  
For spatial data (data frame with an `sf_` prefix) BNG is our default CRS. Some visualisation tools such as leaflet require WGS84, but transformation for this will only be done at the point of visualisation.  
:::

```{r}
#| label: tbl-process
#| tbl-cap: "Data processed (via `_targets.R` & `R\\process.R`)"
#| column: page-right

# Read a list of our processed data from the same spreadsheet
df_processed <- read_xlsx(
  path_transport_data_xls, 
  sheet = "LSOA indicators", 
  skip = 3
)

# List the processed datasets
df_processed |>
  mutate(Row = row_number(), .before = 1) |>
  my_gt() |>
  fmt(
      columns = "Row",
      fns = function(x) {format(round(x, 0), nsmall = 0, big.mark = ",")}
    ) |>
  cols_label(Row = "")
```

The different indicators are assembled into a single data frame. `LSOA21CD` is the key, and there's one row per LSOA and one column per indicator.  

```{r}
#| label: df-lsoa-all

# Load data frame of all LSOA indicators
tar_load(df_lsoa_all)

glimpse(df_lsoa_all)
```

# Handle missing values

```{r}
#| label: tbl-missing
#| tbl-cap: "Missing values by indicator"

df_lsoa_all |>
  summarise(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Indicator", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  mutate(pct_missing = round(n_missing / nrow(df_lsoa_all) * 100, 1)) |>
  arrange(desc(n_missing)) |>
  my_gt() |>
  cols_label(n_missing = "Missing (n)", pct_missing = "Missing (%)")
```

The five DfT traffic indicators (`count_point_n`, `mean_aadf_*`) are missing for `r sum(is.na(df_lsoa_all$count_point_n))` of `r nrow(df_lsoa_all)` LSOAs (78%). DfT count points are placed on strategic roads. Most residential LSOAs simply have no count point within their boundary, so `NA` genuinely means *unmeasured*, not zero.

```{r}
#| label: fig-count-coverage
#| fig-cap: "LSOAs with and without a DfT traffic count point"

tar_load(sf_lsoa)

sf_lsoa |>
  left_join(df_lsoa_all |> select(LSOA21CD, count_point_n), by = "LSOA21CD") |>
  mutate(has_traffic = !is.na(count_point_n)) |>
  ggplot() +
  geom_sf(aes(fill = has_traffic), linewidth = 0.1, colour = "white") +
  scale_fill_manual(
    values = c("TRUE" = "#2166ac", "FALSE" = "#d1d5db"),
    labels = c("TRUE" = "Count point present", "FALSE" = "No count point")
  ) +
  labs(fill = NULL) +
  theme_void()
```

The spatial pattern reflects the arterial road network: LSOAs that contain or border a major road are likely to be covered, while many residential LSOAs are not.

::: {.callout-note}
## Decision  
We're going to run clustering without the DfT indicators. The OS NGD speed metrics and road network characteristics will carry most of the transport environment signal. The DfT indicators would contribute little given their sparse spatial coverage. An alternative is to impute missing traffic values, for example by using the median of the five nearest neighbouring LSOAs (by centroid distance). The DfT targets are retained as they may be reusable if we switch to a spatial denominator larger than LSOAs.   
:::

The single data frame we'll use for clsutering has fewer columns and is called `df_lsoa`.  

```{r}
#| label: df-lsoa

# Load data frame of LSOA indicators we're going to use for clsutering
tar_load(df_lsoa)

glimpse(df_lsoa)
```

# Standardise variables

# Explore distributions
  
  
```{r}
#| label: city-boundary

make_basemap() 
```
  
# Examine correlations

