---
title: "Feature Engineering"
subtitle: "Neighbourhood Transport Environment Typology"
author: "Laurie Platt"
date: last-modified
date-format: "[Last updated ] DD MMMM, YYYY"
format:
  html:
    css: resources/styles.css
    code-tools: true
    self-contained: false
    code-fold: true
    toc: true
    toc-depth: 5
    toc-expand: 1
    number-sections: true
    fig-cap-location: bottom
    fig-align: left
    code-links:
      - text: scc-pi/transport-env-typology
        icon: github
        href: https://github.com/scc-pi/transport-env-typology
execute:
  warning: false
  message: false
---

We're using the {tidymodels}^[{tidymodels} R packages ([www.tidymodels.org](https://www.tidymodels.org/))] framework, a collection of R packages for machine learning. In particular, we're using the {tidyclust}^[{tidyclust} R package ([tidyclust.tidymodels.org](https://tidyclust.tidymodels.org/index.html))] R package.

```{r setup}

# Packages
library(tidyverse)
library(here)
library(targets)
library(gt)
library(gtExtras)
library(sf)
library(leaflet)
library(osdatahub)
library(tidymodels)

# Sources all R/*.R same as _targets.R
tar_source()       

# ggplot theme
set_theme(theme_minimal())
```

The data frame we'll use for clustering is called `df_lsoa` and is a subset of all the indicators we've processed.    

```{r}
#| label: df-lsoa

tar_load(df_lsoa)

glimpse(df_lsoa)
```

# Standardise variables

```{r}
#| label: standardise
#| eval: false 


rec_lsoa <- recipe(~ ., data = df_lsoa) |>

  # LSOA identifier (retain but exclude from modelling)
  update_role(LSOA21CD, new_role = "id") |>

  # Remove zero-variance predictors
  step_zv(all_numeric_predictors()) |>

  # Impute any remaining missing values
  step_impute_median(all_numeric_predictors()) |>

  # Standardise all predictors to mean = 0, sd = 1
  step_normalize(all_numeric_predictors()) |>

  # Remove highly correlated variables
  step_corr(
    all_numeric_predictors(),
    threshold = 0.90
  )

# Prepare recipe
rec_lsoa_prep <- prep(rec_lsoa)

# Create clustering dataset
df_cluster <- bake(
  rec_lsoa_prep,
  new_data = NULL
)

# Predictor matrix for clustering
x_cluster <- df_cluster |>
  select(-LSOA21CD)
```


# Explore distributions
  
  
```{r}
#| label: city-boundary

make_basemap() 
```
  
# Examine correlations

