---
title: "ONS LSOA Classification"
subtitle: "Neighbourhood Transport Environment Typology"
author: "Laurie Platt"
date: last-modified
date-format: "[Last updated ] DD MMMM, YYYY"
format:
  html:
    css: resources/styles.css
    code-tools: true
    self-contained: false
    code-fold: true
    toc: true
    toc-depth: 5
    toc-expand: 1
    number-sections: true
    fig-cap-location: bottom
    fig-align: left
    code-links:
      - text: scc-pi/transport-env-typology
        icon: github
        href: https://github.com/scc-pi/transport-env-typology
execute:
  warning: false
  message: false
---

```{r setup}

# Packages
library(tidyverse)
library(here)
library(targets)
library(gt)
library(gtExtras)
library(sf)
library(leaflet)
library(writexl)

# Sources all R/*.R same as _targets.R
tar_source()       

# ggplot theme
set_theme(theme_minimal())
```

This is a brief exploration of the ONS residential-based LSOA classifications^[ONS 2021 and 2022 residential-based area classifications ([www.ons.gov.uk/methodology/geography/geographicalproducts/areaclassifications/2021and2022residentialbasedareaclassifications#super-output-area-classification-uk-](https://www.ons.gov.uk/methodology/geography/geographicalproducts/areaclassifications/2021and2022residentialbasedareaclassifications#super-output-area-classification-uk-))], hosted by the Geographic Data Service^[GeoDS - UK LSOA / DZ / SDZ Classification (2021/2 LSOAC) ([data.geods.ac.uk/dataset/lsoac](https://data.geods.ac.uk/dataset/lsoac))] (March 2021 to March 2022).  

```{r}
#| label: tbl-oac-etl
#| tbl-cap: "Sample of Sheffield LSOA classifications"

# Sheffield LSOAs already available 
tar_load(sf_lsoa)

# **** Read manual downloads ****

# File with classifications for all UK LSOAs
df_oac_all <- read_csv(
  here(
    "data", "download", "oac", 
    "lsoa_dz_sdz_lookup.csv"
  )
)

# File with names for different classification codes
df_oac_names <- read_csv(
  here(
    "data", "download", "oac", 
    "classification_codes_and_names_ukoac_lsoac.csv")
)

# Only interested in Sheffield LSOA classifications
sf_oac <- sf_lsoa |> 
  left_join(df_oac_all, join_by(LSOA21CD == geography_code))

# **** Work towards a wide data set of classification code and names ****

df_oac_names_subgroup <- df_oac_names |> 
  filter(Level == "Subgroup") |> 
  mutate(
    supergroup_code = str_sub(`Classification Code`, 1, 1),
    group_code = str_sub(`Classification Code`, 1, 2)
  ) |> 
  select(
    supergroup_code,
    group_code,
    subgroup_code = `Classification Code`,
    subgroup_name = `Classification Name`
  )

df_oac_names_group <- df_oac_names |> 
  filter(Level == "Group") |> 
  select(
    group_code = "Classification Code",
    group_name = "Classification Name"
  )

df_oac_names_supergroup <- df_oac_names |> 
  filter(Level == "Supergroup") |> 
  select(
    supergroup_code = "Classification Code",
    supergroup_name = "Classification Name"
  )

df_oac_names_wide <- df_oac_names_subgroup |> 
  left_join(df_oac_names_group, join_by(group_code)) |> 
  left_join(df_oac_names_supergroup, join_by(supergroup_code)) |> 
  relocate(group_name, .after = group_code) |> 
  relocate(supergroup_name, .after = supergroup_code)

# Add the classification names to Sheffield LSOA classifications
sf_oac <- sf_oac |> 
  left_join(df_oac_names_wide, join_by(Subgroup == subgroup_code)) |> 
  rename(subgroup_code = Subgroup) |> 
  select(-Group, -Supergroup)

# No spatial list
df_oac <- sf_oac |> 
  st_drop_geometry() |> 
  select(
    LSOA21CD,
    LSOA21NM,
    supergroup_code,
    supergroup_name,
    group_code,
    group_name,
    subgroup_code,
    subgroup_name
  )

# Export the Sheffield LSOA classifications to a spreadsheet
df_oac |> write_xlsx(here("output", "ons_oac.xlsx"))

# Table of 6 random records
set.seed(123)
df_oac |> 
  slice_sample(n = 6) |> 
  my_gt() |> 
  tab_style(
    style = cell_fill(color = "#D9EAD3"),
    locations = cells_body(columns = 1:2)
  ) |>
  tab_style(
    style = cell_fill(color = "#DCEAF7"),
    locations = cells_body(columns = 3:4)
  ) |>
  tab_style(
    style = cell_fill(color = "#B6D7F0"),
    locations = cells_body(columns = 5:6)
  ) |>
  tab_style(
    style = cell_fill(color = "#6FA8DC"),
    locations = cells_body(columns = 7:8)
  )
```

Pen portraits of the classifications are available in a PDF: [pen_portraits_ukoac_0.pdf](resources/pen_portraits_ukoac_0.pdf) 

```{r}
#| label: fig-map
#| fig-cap: "Sheffield LSOA ONS classifications"
#| column: page
#| fig-height: 6

# Re-project for leaflet
sf_oac <- sf_oac |> 
  st_transform(4326) |> 
  mutate(supergroup_code = as.integer(supergroup_code))

# Palettes
pal_sg  <- colorNumeric(palette = "Oranges", domain = sf_oac$supergroup_code)
pal_grp <- colorFactor(
  palette = colorRampPalette(RColorBrewer::brewer.pal(12, "Paired"))(20),
  domain  = sf_oac$group_code
)

# Bounding box
bb <- sf_oac |> st_bbox()

# Legend label lookups
sg_lookup <- sf_oac |>
  st_drop_geometry() |>
  distinct(supergroup_code, supergroup_name) |>
  arrange(supergroup_code) |>
  mutate(label = paste(supergroup_code, supergroup_name, sep = ". "))

grp_lookup <- sf_oac |>
  st_drop_geometry() |>
  distinct(group_code, group_name) |>
  arrange(group_code) |>
  mutate(label = paste(group_code, group_name, sep = " - "))

# Shared popup
popup_html <- ~paste0(
  "<b>", LSOA21CD, "</b> ", LSOA21NM, "<br>",
  "<b>Supergroup:</b> ", supergroup_code, " - ", supergroup_name, "<br>",
  "<b>Group:</b> ", group_code, " - ", group_name, "<br>",
  "<b>Subgroup:</b> ", subgroup_code, " - ", subgroup_name
)

# Leaflet map
make_basemap() |>
  fitBounds(bb[["xmin"]], bb[["ymin"]], bb[["xmax"]], bb[["ymax"]]) |>
  addPolygons(
    data         = sf_oac,
    group        = "Supergroup",
    smoothFactor = 0.2,
    fillOpacity  = 0.5,
    fillColor    = ~pal_sg(supergroup_code),
    color        = "black",
    weight       = 0.4,
    popup        = popup_html
  ) |>
  addPolygons(
    data         = sf_oac,
    group        = "Group",
    smoothFactor = 0.2,
    fillOpacity  = 0.5,
    fillColor    = ~pal_grp(group_code),
    color        = "black",
    weight       = 0.4,
    popup        = popup_html
  ) |>
  addLayersControl(
    baseGroups = c("Supergroup", "Group"),
    options    = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup("Group") |>
  addLegend(
    position  = "topright",
    colors    = pal_sg(sg_lookup$supergroup_code),
    labels    = sg_lookup$label,
    opacity   = 0.6,
    title     = "Supergroup",
    className = "info legend legend-supergroup"
  ) |>
  addLegend(
    position  = "topright",
    colors    = pal_grp(grp_lookup$group_code),
    labels    = grp_lookup$label,
    opacity   = 0.6,
    title     = "Group",
    className = "info legend legend-group"
  ) |>
  htmlwidgets::onRender(sprintf("
    function(el, x) {
      var map = this;
      // Re-fit bounds with right padding to keep polygons clear of the legend
      map.fitBounds([[%f, %f], [%f, %f]], {paddingBottomRight: [260, 0]});
      // Hide group legend on load
      var grpLegend = document.querySelector('.legend-group');
      if (grpLegend) grpLegend.style.display = 'none';
      // Toggle legends when the base layer changes
      map.on('baselayerchange', function(e) {
        var sgLegend  = document.querySelector('.legend-supergroup');
        var grpLegend = document.querySelector('.legend-group');
        if (e.name === 'Supergroup') {
          if (sgLegend)  sgLegend.style.display  = 'block';
          if (grpLegend) grpLegend.style.display = 'none';
        } else {
          if (sgLegend)  sgLegend.style.display  = 'none';
          if (grpLegend) grpLegend.style.display = 'block';
        }
      });
    }
  ", bb[["ymin"]], bb[["xmin"]], bb[["ymax"]], bb[["xmax"]]))
```

```{r}
#| label: oac-xlsx-gpkg

# Export the Sheffield LSOA classifications to a spreadsheet
df_oac |> write_xlsx(here("output", "ons_oac_sheffield.xlsx"))

# And a geopackage
sf_oac |> st_write(
  dsn = "output/ons_oac_sheffield.gpkg", 
  layer="ons_oac", 
  delete_layer = TRUE, 
  quiet = TRUE
)
```

The data filtered for Sheffield LSOA is available as a spreadsheet and a geopackage:  

* [ons_oac_sheffield.xlsx](output/ons_oac_sheffield.xlsx)  
* [ons_oac_sheffield.gpkg](output/ons_oac_sheffield.gpkg)   

*Source: Geographic Data Service licensed under the Open Government Licence v3.0.*

Not all ONS classifications are present in Sheffield.  

```{r}
#| label: tbl-oac-distinct
#| tbl-cap: "Number of distinct classifications"

df_oac_distinct_ons <- df_oac_names_wide |> 
  summarise(
    classifications = "All ONS",
    n_supergroup = n_distinct(supergroup_code),
    n_group = n_distinct(group_code),
    n_subgroup = n_distinct(subgroup_code)
  )

df_oac_distinct_sheff <- df_oac |> 
  summarise(
    classifications = "Sheffield LSOAs",
    n_supergroup = n_distinct(supergroup_code),
    n_group = n_distinct(group_code),
    n_subgroup = n_distinct(subgroup_code)
  )

df_oac_distinct_ons |> 
  bind_rows(df_oac_distinct_sheff) |> 
  my_gt() |> 
  fmt(
    columns = where(is.numeric),
    fns = function(x) {format(round(x, 0), nsmall = 0, big.mark = ",")}
  )
```

```{r}
#| label: tbl-oac-group-not-sheff
#| tbl-cap: "Group classification not in Sheffield"

df_oac_names_wide |> 
  anti_join(df_oac, join_by(group_code)) |> 
  distinct(
    group_code, group_name, 
    supergroup_code, supergroup_name
  ) |> 
  my_gt() |> 
  tab_style(
    style = cell_fill(color = "#B6D7F0"),
    locations = cells_body(columns = 1:2)
  ) |>
  tab_style(
    style = cell_fill(color = "#DCEAF7"),
    locations = cells_body(columns = 3:4)
  )
```

```{r}
#| label: tbl-oac-subgroup-not-sheff
#| tbl-cap: "Sub group classifications not in Sheffield"

df_oac_names_wide |> 
  anti_join(df_oac, join_by(subgroup_code)) |> 
  select(
    subgroup_code, subgroup_name, 
    group_code, group_name, 
    supergroup_code, supergroup_name
  ) |> 
  my_gt() |>
  tab_style(
    style = cell_fill(color = "#6FA8DC"),
    locations = cells_body(columns = 1:2)
  ) |>
  tab_style(
    style = cell_fill(color = "#B6D7F0"),
    locations = cells_body(columns = 3:4)
  ) |>
  tab_style(
    style = cell_fill(color = "#DCEAF7"),
    locations = cells_body(columns = 5:6)
  )

```