---
title: "Output Area Classification"
subtitle: "Neighbourhood Transport Environment Typology"
author: "Laurie Platt"
date: last-modified
date-format: "[Last updated ] DD MMMM, YYYY"
format:
  html:
    css: resources/styles.css
    code-tools: true
    self-contained: false
    code-fold: true
    toc: true
    toc-depth: 5
    toc-expand: 1
    number-sections: true
    fig-cap-location: bottom
    fig-align: left
    code-links:
      - text: scc-pi/transport-env-typology
        icon: github
        href: https://github.com/scc-pi/transport-env-typology
execute:
  warning: false
  message: false
---

```{r setup}

# Packages
library(tidyverse)
library(here)
library(targets)
library(gt)
library(gtExtras)
library(sf)
library(leaflet)
library(writexl)

# Sources all R/*.R same as _targets.R
tar_source()       

# ggplot theme
set_theme(theme_minimal())
```


This is a brief exploration of the Output Area Classifications^[ONS 2021 and 2022 residential-based area classifications ([www.ons.gov.uk/methodology/geography/geographicalproducts/areaclassifications/2021and2022residentialbasedareaclassifications#super-output-area-classification-uk-](https://www.ons.gov.uk/methodology/geography/geographicalproducts/areaclassifications/2021and2022residentialbasedareaclassifications#super-output-area-classification-uk-))] (OAC), hosted by the Geographic Data Service^[GeoDS - UK LSOA / DZ / SDZ Classification (2021/2 LSOAC) ([data.geods.ac.uk/dataset/lsoac](https://data.geods.ac.uk/dataset/lsoac))] (March 2021 to March 2022).  

The OAC is "a hierarchical geodemographic classification which identifies neighbourhoods across the UK that share similar characteristics. The taxonomy classifies Output Areas (OAs) into 8 different Supergroups, along with 21 nested Groups and 52 nested Subgroups". The OAC for England and Wales is for each LSOA. It is built using official government statistical estimates and the classification methodology is the subject of an academic paper^[Wyszomierski, J., Longley, P.A., Singleton, A.D., Gale, C. & O’Brien, O. (2024) A neighbourhood Output Area Classification from the 2021 and 2022 UK censuses. *The Geographical Journal*, 190, e12550 ([rgs-ibg.onlinelibrary.wiley.com/doi/10.1111/geoj.12550](https://rgs-ibg.onlinelibrary.wiley.com/doi/10.1111/geoj.12550))]. 

Unlike the Acorn and MOSAIC products, OAC is openly available allowing us to explore how NTET might look. NTET is not expected to be hierarchical and will probably only have around half a dozen classifications. The OAC classifications are arguably more intuitive for Local Government evidence-based policy making. However, the OAC is four years old and a lot of the value of the Acorn and MOSAIC products is in the additional underlying data sources, and currency of that data and their product releases.      

```{r}
#| label: tbl-oac-etl
#| tbl-cap: "Sample of Sheffield LSOA classifications"

# Sheffield LSOAs already available 
tar_load(sf_lsoa)

# **** Read manual downloads ****

# File with classifications for all UK LSOAs
df_oac_all <- read_csv(
  here(
    "data", "download", "oac", 
    "lsoa_dz_sdz_lookup.csv"
  )
)

# File with names for different classification codes
df_oac_names <- read_csv(
  here(
    "data", "download", "oac", 
    "classification_codes_and_names_ukoac_lsoac.csv")
)

# Only interested in Sheffield LSOA classifications
sf_oac <- sf_lsoa |> 
  left_join(df_oac_all, join_by(LSOA21CD == geography_code))

# **** Work towards a wide data set of classification code and names ****

df_oac_names_subgroup <- df_oac_names |> 
  filter(Level == "Subgroup") |> 
  mutate(
    supergroup_code = str_sub(`Classification Code`, 1, 1),
    group_code = str_sub(`Classification Code`, 1, 2)
  ) |> 
  select(
    supergroup_code,
    group_code,
    subgroup_code = `Classification Code`,
    subgroup_name = `Classification Name`
  )

df_oac_names_group <- df_oac_names |> 
  filter(Level == "Group") |> 
  select(
    group_code = "Classification Code",
    group_name = "Classification Name"
  )

df_oac_names_supergroup <- df_oac_names |> 
  filter(Level == "Supergroup") |> 
  select(
    supergroup_code = "Classification Code",
    supergroup_name = "Classification Name"
  )

df_oac_names_wide <- df_oac_names_subgroup |> 
  left_join(df_oac_names_group, join_by(group_code)) |> 
  left_join(df_oac_names_supergroup, join_by(supergroup_code)) |> 
  relocate(group_name, .after = group_code) |> 
  relocate(supergroup_name, .after = supergroup_code)

# Add the classification names to Sheffield LSOA classifications
sf_oac <- sf_oac |> 
  left_join(df_oac_names_wide, join_by(Subgroup == subgroup_code)) |> 
  rename(subgroup_code = Subgroup) |> 
  select(-Group, -Supergroup)

# No spatial list
df_oac <- sf_oac |> 
  st_drop_geometry() |> 
  select(
    LSOA21CD,
    LSOA21NM,
    supergroup_code,
    supergroup_name,
    group_code,
    group_name,
    subgroup_code,
    subgroup_name
  )

# Table of 6 random records
set.seed(123)
df_oac |> 
  slice_sample(n = 6) |> 
  my_gt() |> 
  tab_style(
    style = cell_fill(color = "#D9EAD3"),
    locations = cells_body(columns = 1:2)
  ) |>
  tab_style(
    style = cell_fill(color = "#DCEAF7"),
    locations = cells_body(columns = 3:4)
  ) |>
  tab_style(
    style = cell_fill(color = "#B6D7F0"),
    locations = cells_body(columns = 5:6)
  ) |>
  tab_style(
    style = cell_fill(color = "#6FA8DC"),
    locations = cells_body(columns = 7:8)
  )
```

```{r}
#| label: fig-map
#| fig-cap: "Sheffield OAC"
#| column: page
#| fig-height: 6

# Re-project for leaflet
sf_oac <- sf_oac |> 
  st_transform(4326) |> 
  mutate(supergroup_code = as.integer(supergroup_code))

# Palettes
pal_sg  <- colorNumeric(palette = "Oranges", domain = sf_oac$supergroup_code)
pal_grp <- colorFactor(
  palette = colorRampPalette(RColorBrewer::brewer.pal(12, "Paired"))(20),
  domain  = sf_oac$group_code
)

# Bounding box
bb <- sf_oac |> st_bbox()

# Legend label lookups
sg_lookup <- sf_oac |>
  st_drop_geometry() |>
  distinct(supergroup_code, supergroup_name) |>
  arrange(supergroup_code) |>
  mutate(label = paste(supergroup_code, supergroup_name, sep = ". "))

grp_lookup <- sf_oac |>
  st_drop_geometry() |>
  distinct(group_code, group_name) |>
  arrange(group_code) |>
  mutate(label = paste(group_code, group_name, sep = " - "))

# Shared popup
popup_html <- ~paste0(
  "<b>", LSOA21CD, "</b> ", LSOA21NM, "<br>",
  "<b>Supergroup:</b> ", supergroup_code, " - ", supergroup_name, "<br>",
  "<b>Group:</b> ", group_code, " - ", group_name, "<br>",
  "<b>Subgroup:</b> ", subgroup_code, " - ", subgroup_name
)

# Leaflet map
make_basemap() |>
  fitBounds(bb[["xmin"]], bb[["ymin"]], bb[["xmax"]], bb[["ymax"]]) |>
  addPolygons(
    data         = sf_oac,
    group        = "Supergroup",
    smoothFactor = 0.2,
    fillOpacity  = 0.5,
    fillColor    = ~pal_sg(supergroup_code),
    color        = "black",
    weight       = 0.4,
    popup        = popup_html
  ) |>
  addPolygons(
    data         = sf_oac,
    group        = "Group",
    smoothFactor = 0.2,
    fillOpacity  = 0.5,
    fillColor    = ~pal_grp(group_code),
    color        = "black",
    weight       = 0.4,
    popup        = popup_html
  ) |>
  addLayersControl(
    baseGroups = c("Supergroup", "Group"),
    options    = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup("Group") |>
  addLegend(
    position  = "topright",
    colors    = pal_sg(sg_lookup$supergroup_code),
    labels    = sg_lookup$label,
    opacity   = 0.6,
    title     = "Supergroup",
    className = "info legend legend-supergroup"
  ) |>
  addLegend(
    position  = "topright",
    colors    = pal_grp(grp_lookup$group_code),
    labels    = grp_lookup$label,
    opacity   = 0.6,
    title     = "Group",
    className = "info legend legend-group"
  ) |>
  htmlwidgets::onRender(sprintf("
    function(el, x) {
      var map = this;
      // Re-fit bounds with right padding to keep polygons clear of the legend
      map.fitBounds([[%f, %f], [%f, %f]], {paddingBottomRight: [260, 0]});
      // Hide group legend on load
      var grpLegend = document.querySelector('.legend-group');
      if (grpLegend) grpLegend.style.display = 'none';
      // Toggle legends when the base layer changes
      map.on('baselayerchange', function(e) {
        var sgLegend  = document.querySelector('.legend-supergroup');
        var grpLegend = document.querySelector('.legend-group');
        if (e.name === 'Supergroup') {
          if (sgLegend)  sgLegend.style.display  = 'block';
          if (grpLegend) grpLegend.style.display = 'none';
        } else {
          if (sgLegend)  sgLegend.style.display  = 'none';
          if (grpLegend) grpLegend.style.display = 'block';
        }
      });
    }
  ", bb[["ymin"]], bb[["xmin"]], bb[["ymax"]], bb[["xmax"]]))
```

Pen portraits of all the classifications are available in a PDF: [pen_portraits_ukoac_0.pdf](resources/pen_portraits_ukoac_0.pdf). Here are the pen portraits for the Supergroups:    

1. **Retired Professionals**. Typically married but no longer with resident dependent children, these well-educated households either remain working in their managerial, professional, administrative or other skilled occupations, or are retired from them – the modal individual age is beyond normal retirement age. Detached and semi-detached properties predominate, and unpaid care is more prevalent than reported disability. The prevalence of this Supergroup outside most urban conurbations indicates that rural lifestyles prevail, typically sustained by using two or more cars per household.  
1. **Suburbanites and Peri-Urbanites**. Pervasive throughout the UK, members of this Supergroup typically own (or are buying) their detached, semi-detached or terraced homes. They are also typically educated to A Level/Highers or degree level and work in skilled or professional or management occupations. Typically born in the UK, some families have children, although the median adult age is above 45. This Supergroup is pervasive not only in suburban locations, but also in neighbourhoods at or beyond the edge of cities that adjoin rural parts of the country.  
1. **Multicultural and Educated Urbanites**. Established populations comprising ethnic minorities together with persons born outside the UK predominate in this Supergroup. Residents present diverse personal characteristics and circumstances: while generally well-educated and practising skilled occupations, there is low levels of home ownership. English may not be the main language used by people in this Group. Although the typical adult resident is middle aged, single person households are common and marriage rates are low by national standards. This Supergroup predominates in Inner London, with smaller enclaves in many other densely populated metropolitan areas.   
1. **Low-Skilled Migrant and Student Communities**. Young adults, many of whom are students, predominate in these high-density neighbourhoods of rented terrace houses or flats. Most ethnic minorities are present in these communities, as are people born outside of the UK. Students aside, low skilled occupations predominate, and unemployment rates are above average. The Mixed or Multiple ethnic group composition of neighbourhoods is often associated with low rates of affiliation to Christian religions. This Supergroup predominates in non-central urban locations across Britain, particularly in the Midlands and the outskirts of west, south and north-east London.  
1. **Diverse Suburban Professionals**. Those working within the managerial, professional and administrative occupations typically reflect a wide range of ethnic groups, and reside in detached or semi-detached housing. Their residential locations at the edges of cities and conurbations and car-based lifestyles are more characteristic of Supergroup membership than birthplace or participation in child-rearing. Houses are typically owner-occupied and marriage rates are higher than the national average. This Supergroup is found throughout suburban Britain.  
1. **Baseline UK**. This Supergroup exemplifies the broad base to the UK’s social structure, encompassing as it does the average or modal levels of many neighbourhood characteristics, including all housing tenures, a range of levels of educational attainment and religious affiliations, and a variety of pre-retirement age structures. Yet, in combination, these mixes are each distinctive of the parts of the UK. Overall, terraced houses are the most prevalent, as is employment in intermediate or low-skilled occupations. However, this Supergroup is also characterised by above average levels of unemployment and lower levels of use of English as the main language. Many neighbourhoods occur in south London and the UK’s other major urban centres.  
1. **Semi- and Un-Skilled Workforce**. Living in terraced or semi-detached houses, residents of these neighbourhoods typically lack high levels of education and work in elementary or routine service occupations. Unemployment is above average. Residents are predominantly born in the UK. Social (but not private sector) rented sector housing is common. This Supergroup is found throughout the UK’s conurbations and industrial regions but is also an integral part of smaller towns.  
1. **Legacy Communities**. These neighbourhoods characteristically comprise pockets of flats that are scattered across the UK, particularly in towns that retain or have legacies of heavy industry or are in more remote seaside locations. Employed residents of these neighbourhoods work mainly in low-skilled occupations. Residents typically have limited educational qualifications. Unemployment is above average. Many residents live in the social rented sector and experience long-term disability. All adult age groups are represented, although there is an overall age bias towards elderly people in general and the very old in particular.  

```{r}
#| label: fig-sheff-oac
#| fig-cap: "OAC Supergroup classifications in Sheffield by number of LSOAs"

df_oac |>
  count(supergroup_code, supergroup_name) |>
  mutate(supergroup_label = str_c(supergroup_code, ". ", supergroup_name)) |> 
  mutate(supergroup_label = fct_reorder(supergroup_label, n)) |>
  ggplot(aes(x = n, y = supergroup_label)) +
  geom_col() +
  labs(
    x = "Number of LSOAs",
    y = NULL
  )
```


```{r}
#| label: oac-xlsx-gpkg

# Export the Sheffield LSOA classifications to a spreadsheet
df_oac |> write_xlsx(here("output", "ons_oac_sheffield.xlsx"))

# And a geopackage
sf_oac |> st_write(
  dsn = "output/ons_oac_sheffield.gpkg", 
  layer="ons_oac", 
  delete_layer = TRUE, 
  quiet = TRUE
)
```

The data filtered for Sheffield LSOA is available as a spreadsheet and a geopackage:  

* [ons_oac_sheffield.xlsx](output/ons_oac_sheffield.xlsx)  
* [ons_oac_sheffield.gpkg](output/ons_oac_sheffield.gpkg)   

*Source: Geographic Data Service licensed under the Open Government Licence v3.0.*

Not all ONS classifications are present in Sheffield.  

```{r}
#| label: tbl-oac-distinct
#| tbl-cap: "Number of distinct classifications"

df_oac_distinct_ons <- df_oac_names_wide |> 
  summarise(
    classifications = "All ONS",
    n_supergroup = n_distinct(supergroup_code),
    n_group = n_distinct(group_code),
    n_subgroup = n_distinct(subgroup_code)
  )

df_oac_distinct_sheff <- df_oac |> 
  summarise(
    classifications = "Sheffield LSOAs",
    n_supergroup = n_distinct(supergroup_code),
    n_group = n_distinct(group_code),
    n_subgroup = n_distinct(subgroup_code)
  )

df_oac_distinct_ons |> 
  bind_rows(df_oac_distinct_sheff) |> 
  my_gt() |> 
  fmt(
    columns = where(is.numeric),
    fns = function(x) {format(round(x, 0), nsmall = 0, big.mark = ",")}
  )
```

```{r}
#| label: tbl-oac-group-not-sheff
#| tbl-cap: "Group classification not in Sheffield"

df_oac_names_wide |> 
  anti_join(df_oac, join_by(group_code)) |> 
  distinct(
    group_code, group_name, 
    supergroup_code, supergroup_name
  ) |> 
  my_gt() |> 
  tab_style(
    style = cell_fill(color = "#B6D7F0"),
    locations = cells_body(columns = 1:2)
  ) |>
  tab_style(
    style = cell_fill(color = "#DCEAF7"),
    locations = cells_body(columns = 3:4)
  )
```

```{r}
#| label: tbl-oac-subgroup-not-sheff
#| tbl-cap: "Sub group classifications not in Sheffield"

df_oac_names_wide |> 
  anti_join(df_oac, join_by(subgroup_code)) |> 
  select(
    subgroup_code, subgroup_name, 
    group_code, group_name, 
    supergroup_code, supergroup_name
  ) |> 
  my_gt() |>
  tab_style(
    style = cell_fill(color = "#6FA8DC"),
    locations = cells_body(columns = 1:2)
  ) |>
  tab_style(
    style = cell_fill(color = "#B6D7F0"),
    locations = cells_body(columns = 3:4)
  ) |>
  tab_style(
    style = cell_fill(color = "#DCEAF7"),
    locations = cells_body(columns = 5:6)
  )

```